P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2024-01-09 manuscript.tex
minor
Signed-off-by: Rafael Ravedutti <rafael.r.ravedutti@fau.de>

\autoref{fig:md_single_core_runtimes} display the runtime measurements for P4IRS and MD-Bench simulations of the Copper FCC lattice case with $32^{3}$ unit cells (4 atoms per unit-cell).
It is noticeable that the kernels generated by P4IRS achieve competitive results in comparison to MD-Bench, a hand-tuned implementation for MD cases, thus demonstrating P4IRS capabilities to generate fast code.
For Ice Lake, performance is slightly better for MD-Bench using \ac{FN}, with about 10\% speedup for the force calculation but 25\% worse performance for the neighbor-lists creation.
When using \ac{HN}, the force calculation time is roughly the same for both versions, but P4IRS still keeps the advantage when bulding the neighbor-lists, resulting in a 4\% better performance for P4IRS in the overall runtime.
Note that the code from P4IRS tends to be build neighbor-lists faster for these experiments but the force computation is slower for \ac{FN}, and the runtime differences vary in the range of 10\% for the measured runtime, which is not too significant and can vary when using different compilers and flags.

@@

\autoref{fig:md_single_gpu_runtimes} displays the runtime measurements for P4IRS and MD-Bench simulations in the A40 and A100 GPUs.
MD-Bench still do not have a GPU version with \ac{HN}, so we just display it using \ac{FN}.
For A40, MD-Bench shows a slightly better performance (about 3\%) for computing the forces while P4IRS with \ac{FN} also provides slighly better performance (about 7\%) for building the neighbor-lists.
In the overall performance, MD-Bench is faster only by about 2\% in comparison to P4IRS for the \ac{FN} case, which is not too significant and corroborates our claim that P4IRS is able to generate efficient GPU code.
For the \ac{HN} case, the performance is not even close to a factor of two as one would expect, which can be attributed to the usage of atomics, and this can be different for more arithmetic-intensive kernels (the \ac{LJ} kernel used has significantly less arithmetic instructions compared to realistic cases) since the overall amount of operations reduced should be larger.
For A100, however, the performance difference is larger, where MD-Bench performance is about 14\% faster in comparison to P4IRS.
When using the \ac{HN}, however, we achieve very close performance to MD-Bench with \ac{FN}, however it is noticeable that the speedup is very far from the factor of two as well.
Despite the observed performance divergence in A100, it is possible to see that P4IRS code still has good performance in comparison to our hand-tuned CUDA implementation, supporting our claim in generating performance-portable kernels.

@@

\autoref{fig:md_weak_scaling_juwels} shows the weak-scaling runtimes for P4IRS in the JUWELS Booster cluster.
Each node simulates a system of $100^3$ unit cells, with roughly 500.000 atoms per GPU.
We scale up to 128 nodes, which is equivalent of 512 A100 GPUs.
Note that for the first node, the performance is slightly worse than the next ones, which can be explained as we only have 4 GPUs in one node, and domain-partitioning is only done in two dimensions.
This can led to worse performance because the surface area for communication is likely higher for the dimension where the partitioning is not done, which culminates in more atoms to be communicated and therefore more time spent packing and transfering data.
Apart from this, the scaling is also perfect for all the settings, and it is possible to observe that the \ac{FN} version is slighly faster than \ac{HN}.
As we already discussed, the usage of \ac{HN} is not necessarily beneficial in GPUs due to atomics, and this can be worse especially for kernels which do not exhibit high arithmetic intensity.
The results demonstrate that P4IRS is in fact capable of generating scalable code for modern GPU clusters.

@@

\autoref{fig:dem_single_gpu_runtimes} displays the runtimes for the DEM testcase in both A40 and A100 GPUs, also evaluating the neighbor-lists per cell (see \autoref{sec:neighbor_lists_per_cell}) and multi-kernel (see \autoref{sec:shape_partitioning}) strategies.
Most of the simulation time is spent on calculating the forces and executing other routines, which is likely transferring data from and to the GPUs.
Note that since particles contain significant more properties (and some of them with dynamic types such as vectors and matrices), the communication and/or PBC buffers are significantly larger especially when particles need to be exchanged.
It is also important to disclose that due to PBC particles are still exchanged and produce ghost versions of themselves (with proper PBC shifts) even when running in a single MPI rank, therefore the same routines for packing and unpacking data for communication are used.

@@

Results for \ac{MD} kernels generated by P4IRS demonstrate that we can achieve state-of-the-art performance in comparison to our performance-focused prototyping harness MD-Bench, which contains hand-tuned implementations of \ac{MD} kernels in C and CUDA.
Also, results in modern clusters Fritz and JUWELS Booster are displayed to show P4IRS scaling capabilities, which are important for particle simulations in general.
For more complex and realistic cases, we also show that P4IRS is able to generate code to solve \ac{DEM} simulation with the Linear Spring-Dashpot contact model.
The performance results demonstrate that P4IRS significantly outperforms our previous framework (called MESA-PD), which can be attributed to aggresive specializations of the force calculation kernels for the contact model.
Besides, we also demonstrate how P4IRS can be used for experimentation, since contact models or potentials can be generated with different optimization strategies without the effort of implementing new kernels, highlighting the advantages of using code generation techniques.

In the future, we intend to use P4IRS particle simulations coupled with fluid simulations, which can be solved through the waLBerla framework.
For this, we still need to generate proper interfaces for the coupling and generate modular code for the particle simulations that can be seamlessly integrated into other applications.
Nevertheless, we also want to include more performance optimizations into P4IRS such as the Cluster Pair algorithm from GROMACS to enhance SIMD performance, as well as being able to generate SIMD intrinsics explicitly into P4IRS generated kernels to either enforce SIMD vectorization and also to be able to choose between different strategies (such as not using gather instructions explicitly).
At the end, the separation of concerns introduced by P4IRS together with its intermediate representation embedded in Python provides several opportunities to explore and experiment with performance in different types of particle simulations.

>>>>

\autoref{fig:md_single_core_runtimes} display the runtime measurements for P4IRS and MD-Bench simulations of the Copper FCC lattice case with $32^{3}$ unit cells (4 atoms per unit-cell).
It is noticeable that the kernels generated by P4IRS achieve competitive results in comparison to MD-Bench, a hand-tuned implementation for MD cases, thus demonstrating P4IRS capabilities to generate fast code.
For Ice Lake, performance is slightly better for MD-Bench using \ac{FN}, with about 10\% speedup for the force calculation but 25\% worse performance for the neighbor-lists creation.
When using \ac{HN}, the force calculation time is roughly the same for both versions, but P4IRS still keeps the advantage when bulding the neighbor-lists, resulting in a 4\% better performance for P4IRS in the overall runtime.
Overall, differences vary in the range of 10\% (at maximum) of the measured runtime, which is not too significant and can be different when using different compilers and flags.

@@

\autoref{fig:md_single_gpu_runtimes} displays the runtime measurements for P4IRS and MD-Bench simulations in the A40 and A100 GPUs.
MD-Bench still do not have a GPU version with \ac{HN}, so we just display it using \ac{FN}.
For A40, MD-Bench shows a slightly better performance (about 3\%) for computing the forces while P4IRS with \ac{FN} also provides slighly better performance (about 7\%) for building the neighbor-lists.
In the overall performance, MD-Bench is faster only by about 2\% in comparison to P4IRS for the \ac{FN} case, which is not too significant and corroborates our claim that P4IRS is able to generate efficient GPU code.
For the \ac{HN} case, the performance is not even close to a factor of two as one would expect, which can be attributed to the usage of atomics, and this can be different for more arithmetic-intensive kernels (the \ac{LJ} kernel used has significantly less arithmetic instructions compared to realistic cases) since the overall amount of operations reduced should be larger.
For A100, however, the performance difference is larger, where MD-Bench performance is about 14\% faster in comparison to P4IRS.
When using the \ac{HN}, however, we achieve very close performance to MD-Bench with \ac{FN}, and it is noticeable that the speedup is very far from the factor of two as well.
Despite the observed performance divergence in A100, it is possible to see that P4IRS code still has good performance in comparison to our hand-tuned CUDA implementation, supporting our claim in generating performance-portable kernels.

@@

\autoref{fig:md_weak_scaling_juwels} shows the weak-scaling runtimes for P4IRS in the JUWELS Booster cluster.
Each node simulates a system of $100^3$ unit cells, with roughly 500.000 atoms per GPU.
We scale up to 128 nodes, which is equivalent of 512 A100 GPUs.
Note that for the first node, the performance is slightly worse than the next ones, which can be explained as we only have 4 GPUs in one node, and domain-partitioning is only done in two dimensions.
This can lead to worse performance because the surface area for communication is likely higher for the dimension where the partitioning is not done, which culminates in more atoms to be communicated and therefore more time spent packing and transfering data.
Apart from this, the scaling is also perfect for all the settings, and it is possible to observe that the \ac{FN} version is slighly faster than \ac{HN}.
As we already discussed, the usage of \ac{HN} is not necessarily beneficial in GPUs due to atomics, and this can be worse especially for kernels which do not exhibit high arithmetic intensity.
The results demonstrate that P4IRS is in fact capable of generating scalable code for modern GPU clusters.

@@

\autoref{fig:dem_single_gpu_runtimes} displays the runtimes for the DEM testcase in both A40 and A100 GPUs, also evaluating the neighbor-lists per cell (see \autoref{sec:neighbor_lists_per_cell}) and multi-kernel (see \autoref{sec:shape_partitioning}) strategies.
Most of the simulation time is spent on calculating the forces and executing other routines, which is likely transferring data into/from the GPUs.
Note that since particles contain significant more properties (and some of them with dynamic types such as vectors and matrices), the communication and/or PBC buffers are significantly larger especially when particles need to be exchanged.
It is also important to disclose that due to PBC particles are still exchanged and produce ghost versions of themselves (with proper PBC shifts) even when running in a single MPI rank, therefore the same routines for packing and unpacking data for communication are used.

@@

Results for \ac{MD} kernels generated by P4IRS demonstrate that we can achieve state-of-the-art performance in comparison to our performance-focused prototyping harness MD-Bench, which contains hand-tuned implementations of \ac{MD} kernels in C and CUDA.
Also, results in modern clusters Fritz and JUWELS Booster are displayed to show P4IRS scaling capabilities, which are important for particle simulations in general.
For more complex and realistic cases, we also show that P4IRS is able to generate code to solve \ac{DEM} simulation with the Linear Spring-Dashpot contact model.
The performance results demonstrate that P4IRS significantly outperforms our previous framework (called MESA-PD), which can be attributed to aggresive specializations in the force calculation kernels of the contact model.
Besides, we also demonstrate how P4IRS can be used for experimentation, since contact models or potentials can be generated with different optimization strategies without the effort of implementing new kernels, highlighting the advantages of using code generation techniques.

In the future, we intend to use P4IRS particle simulations coupled with fluid simulations, which can be solved through the waLBerla framework.
For this, we still need to generate proper interfaces for the coupling and generate modular code for the particle simulations that can be seamlessly integrated into other applications.
Nevertheless, we also want to include more performance optimizations into P4IRS such as the Cluster Pair algorithm from GROMACS to enhance SIMD performance, as well as being able to generate SIMD intrinsics explicitly into P4IRS generated kernels to either enforce SIMD vectorization and also to be able to choose between different strategies (i.e. gather data manually instead of using gather instructions).
At the end, the separation of concerns introduced by P4IRS together with its intermediate representation embedded in Python provide several opportunities to explore and experiment with performance on many types of particle simulations that can belong to distinct research fields.
